A tool set for random number generation on GPUs in R

### 2021-12-26 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3069 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
Random number generation is more complicated when RNG is used in parallel processes in \proglang{R}, there is a risk that the generated sequences of random numbers have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it.

In the \proglang{clRNG} library, the MRG31k3p RNG's entire period of length approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. %Each stream can be further partitioned into substreams of length $2^{72}$. 
The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed for the package'' in \proglang{clRNG} and ``initial'' in \pkg{clrng}) for the MRG31k3p is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

@@

\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: vector of length 6, recycled if shorter.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'. \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large. However, any data is lost when the \proglang{R} session is terminated. %\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
\begin{CodeChunk}
\begin{CodeInput}
R> # creating streams on GPU
R> myStreamsGpu = vclMatrix(myStreamsCpu)
R> myStreamsGpu2 = createStreamsGpu(n = 4, initial = 12345)
\end{CodeInput} 
\end{CodeChunk}

@@

Now we use the streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with \fct{clrng::runif}. To view the generated random numbers, we need to convert them to \proglang{R} vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{CodeChunk}
\begin{CodeInput}
R> as.vector(clrng::runif(n = 6, streams = myStreamsGpu, Nglobal = c(2,
+    2)))
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.735 0.842 0.614 0.216 0.110 0.870
\end{CodeOutput} 
\end{CodeChunk}

@@

Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is the code that saves streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
\begin{CodeChunk}
\begin{CodeInput}
R> saveRDS(as.matrix(createStreamsGpu(n = 4)), "myStreams.rds")
R> # Load the streams object as streams_saved
R> streams_saved <- vclMatrix(readRDS("myStreams.rds"))
\end{CodeInput} 
\end{CodeChunk}

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.048   0.004   0.051
\end{CodeOutput} 
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> system.time(matrix(stats::rnorm(10000^2), 10000, 10000))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   6.978   0.847   7.821
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a $2 \times 4$ matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeInput}
R> r_matrix <- clrng::rexp(c(2, 4), rate = 1, Nglobal = c(2,
+    2), type = "double")
R> as.matrix(r_matrix)
\end{CodeInput}
\begin{CodeOutput}
##       [,1]  [,2]  [,3]   [,4]
## [1,] 0.565 0.333 0.395 0.0932
## [2,] 1.535 0.866 0.124 0.0315
\end{CodeOutput} 
\end{CodeChunk}

The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64, initial = 666)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.311   0.005   0.315
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 410825
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.404
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_monthcpu <- stats::fisher.test(month, simulate.p.value = TRUE,
+    B = 1015808))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  14.791   0.027  14.826
\end{CodeOutput}
\begin{CodeInput}
R> result_monthcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.405
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.931   0.023   1.953
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1193
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000119
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_weekcpu <- fisher.test(week, simulate.p.value = TRUE,
+    B = 10010624))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  92.369   0.143  92.455
\end{CodeOutput}
\begin{CodeInput}
R> result_weekcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000127
\end{CodeOutput} 
\end{CodeChunk}

\subsection{A summary of the results}
We summarized the comparison results in Table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
\begin{table}

@@

\begin{figure}

>>>>

%THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.
\subsection{Creating streams}\label{createstreams}
When a RNG is called in parallel processes or successively called at several places in a program in \proglang{R}, random number generation would be more complicated because there is no guarantee that the streams (or similarly \code{.Random.seed} in \proglang{R}) do not overlap, and thus the generated sequences of random numbers is possible to have correlation. \pkg{clrng} uses multiple distinct streams that are used in work-items that executes in parallel on a GPU device. A popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long disjoint pieces of equal length $Z$, and use each piece as a separate stream. Creating a new stream amounts to computing its starting point. Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, although this is not currently implemented in \pkg{clrng}. In general, a stream object contains three elements: the current state of the stream, the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Streams are created sequentially in the way that whenever the user creates a new stream, the software automatically jumps ahead by $Z$ steps to find its initial state, and the three states in the stream object are set to it.

In the \proglang{clRNG} library, the MRG31k3p RNG's entire period of length approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. %Each stream can be further partitioned into substreams of length $2^{72}$. 
The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, while providing a sufficient period length for most applications. The initial state of the first stream (also called ``initial seed of the package'' or ``initial seed of the stream creator'' in \proglang{clRNG}) for the MRG31k3p RNG is by default $(12345, 12345, 12345, 12345, 12345, 12345)$.

@@

\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: a vector of six 31-bit integers, recycled if length is shorter. Default is set to \code{c(12345, 12345, 12345, 12345, 12345, 12345)}.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'. \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large. Users may choose either \fct{createStreamsCpu} or \fct{createStreamsGpu} and stick with that choice for their entire program, as we want to avoid the very small possibility of overlap between the streams with different initial seeds. The `\code{initial}' argument should be set only once at the very beginning of the program in each \proglang{R} session if users want to supply their own initial seed of the stream creator, or never, in which case users would go with the package default initial seed. After the `\code{initial}' argument is set, we can ignore the `\code{initial}' argument and call \fct{createStreamsCpu} or \fct{createStreamsGpu} once or several times to create various matrices of streams in the same program.

%However, any data is lost when the \proglang{R} session is terminated. %\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
\begin{CodeChunk}
\begin{CodeInput}
R> # creating streams on GPU
R> myStreamsGpu = vclMatrix(myStreamsCpu)
\end{CodeInput} 
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> myStreamsGpu2 = createStreamsGpu(n = 4, initial = 12345)
\end{CodeInput} 
\end{CodeChunk}

@@

Now we use the 4 streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with \fct{clrng::runif}. To view the generated random numbers, we need to convert them to \proglang{R} vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{CodeChunk}
\begin{CodeInput}
R> as.vector(clrng::runif(n = 6, streams = myStreamsGpu2, Nglobal = c(2,
+    2)))
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.735 0.842 0.614 0.216 0.110 0.870
\end{CodeOutput} 
\end{CodeChunk}

@@

Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams' current states and initial states in matrix form on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, the streams will restart from the current states after they are reloaded. In this way we can reproduce the exact same sequence of random numbers in simulations. Below is the code that saves streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
\begin{CodeChunk}
\begin{CodeInput}
R> saveRDS(as.matrix(createStreamsGpu(n = 4)), "myStreams.rds")
R> # Load the streams object as streams_saved
R> streams_saved <- vclMatrix(readRDS("myStreams.rds"))
\end{CodeInput} 
\end{CodeChunk}

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.043   0.007   0.050
\end{CodeOutput} 
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> system.time(matrix(stats::rnorm(10000^2), 10000, 10000))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   6.947   0.736   7.686
\end{CodeOutput} 
\end{CodeChunk}

@@

Below is an example that produces a $2 \times 4$ matrix of Exponential random numbers with expectation equal to 1.
\begin{CodeChunk}
\begin{CodeInput}
R> r_matrix <- clrng::rexp(c(2, 4), rate = 1, Nglobal = c(2,
+    2), type = "double")
R> as.matrix(r_matrix)
\end{CodeInput}
\begin{CodeOutput}
##        [,1]  [,2]  [,3]  [,4]
## [1,] 0.1187 0.187 1.079 0.112
## [2,] 0.0501 0.369 0.411 1.880
\end{CodeOutput} 
\end{CodeChunk}

The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, streams created on GPU with package default initial seed are supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.318   0.005   0.322
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 409435
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.403
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_monthcpu <- stats::fisher.test(month, simulate.p.value = TRUE,
+    B = 1015808))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  14.859   0.014  14.874
\end{CodeOutput}
\begin{CodeInput}
R> result_monthcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.405
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##    1.93    0.02    1.95
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1351
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000135
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # using CPU
R> system.time(result_weekcpu <- fisher.test(week, simulate.p.value = TRUE,
+    B = 10010624))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##  91.311   0.186  91.432
\end{CodeOutput}
\begin{CodeInput}
R> result_weekcpu$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000127
\end{CodeOutput} 
\end{CodeChunk}

\subsection{A summary of the results}
We summarized the comparison results in Table \ref{tab:summary} and plot the test statistics in Figure \ref{fig4}.
\begin{table}[H]

@@

\begin{figure}[H]
